---
title: "Coursera Case Study"
author: "UnYoung"
date: '2023-03-14'
output: html_document
---

# Cyclicist 12-Month Analysis

Cyclistic, a bike-share company in Chicago, wants to understand how casual riders and annual members use Cyclistic bikes differently. The question at hand is: How do annual members and casual riders use Cyclistic bikes differently?

## Data preparation

This describes the data sources used and documents data cleaning and manipulation.

-   Data source: Cyclistic's historical trip data

-   Time period: March 2022 to February 2023

### Data cleaning

-   Download csv files ("202203-divvy-tripdata.csv", "202204-divvy-tripdata.csv", "202205-divvy-tripdata.csv", "202206-divvy-tripdata.csv","202207-divvy-tripdata.csv","202208-divvy-tripdata.csv","202209-divvy-publictripdata.csv","202210-divvy-tripdata.csv","202211-divvy-tripdata.csv","202212-divvy-tripdata.csv","202301-divvy-tripdata.csv","202302-divvy-tripdata.csv") and load into R
-   Compare column names each of the files
-   Merge dataframes
-   Keep only necessary columns
-   Inspect merged dataframe
-   Drop incorrect or missing data

```{r data prep, include=FALSE}
#Load packages
library(readr)
library(tidyverse)
library(lubridate)
library(ggplot2) 

#Load data
# Upload Divvy datasets (csv files) here
m3_2022 <- read_csv("202203-divvy-tripdata.csv")
m4_2022 <- read_csv("202204-divvy-tripdata.csv")
m5_2022 <- read_csv("202205-divvy-tripdata.csv")
m6_2022 <- read_csv("202206-divvy-tripdata.csv")
m7_2022 <- read_csv("202207-divvy-tripdata.csv")
m8_2022 <- read_csv("202208-divvy-tripdata.csv")
m9_2022 <- read_csv("202209-divvy-publictripdata.csv")
m10_2022 <- read_csv("202210-divvy-tripdata.csv")
m11_2022 <- read_csv("202211-divvy-tripdata.csv")
m12_2022 <- read_csv("202212-divvy-tripdata.csv")
m1_2023 <- read_csv("202301-divvy-tripdata.csv")
m2_2023 <- read_csv("202302-divvy-tripdata.csv")

#Check column names --> Column names are consistent throughout the dataframes
names(m3_2022)
names(m4_2022)
names(m5_2022)
names(m6_2022)
names(m7_2022)
names(m8_2022)
names(m9_2022)
names(m10_2022)
names(m11_2022)
names(m12_2022)
names(m1_2023)
names(m2_2023)

#Initial inspection of each dataframe
glimpse(m3_2022)
glimpse(m4_2022)
glimpse(m5_2022)
glimpse(m6_2022)
glimpse(m7_2022)
glimpse(m8_2022)
glimpse(m9_2022)
glimpse(m10_2022)
glimpse(m11_2022)
glimpse(m12_2022)
glimpse(m1_2023)
glimpse(m2_2023)

#Merge dataframes
all_trips <- bind_rows(m3_2022, m4_2022, m5_2022, m6_2022, m7_2022, m8_2022, m9_2022, m10_2022, m11_2022, m12_2022, m1_2023, m2_2023)

all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))

#Inspect dataframe
table(all_trips$member_casual)

#Add dates
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

#Add ride lengths
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

glimpse(all_trips)

is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

#Remove incorrect or missing data
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<=0),]
all_trips_v2 <- na.omit(all_trips_v2)
```

## Analysis

```{r analysis, echo=TRUE, message=FALSE, warning=FALSE}
# Descriptive analysis
summary(all_trips_v2$ride_length)

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("maandag", "dinsdag", "woensdag", "donderdag", "vrijdag", "zaterdag", "zondag"))

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

# Ridership data by type and weekday
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n(),							#calculates the number of rides and average duration 
  average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday) %>%
  summary()


```

## Visualizations

```{r viz, echo=FALSE, message=FALSE, warning=FALSE}
numrides <- all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "Number of Rides", fill = "Membership Type") + ylab("Number of Rides") + xlab("Weekday") + 
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000), labels=c("0", "100k", "200k", "300k", "400k")) +
  scale_x_discrete(labels=c("Sun", "Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun"))

plot(numrides)

duration <- all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "Average Duration of Rides", fill = "Membership Type") + ylab("Duration in Seconds") + xlab("Weekday") +
  scale_x_discrete(labels=c("Sun", "Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun"))

plot(duration)
```

## Results

-   Members have a higher number of rides during the week compared to casual riders.
-   Members and casual riders have similar number of rides on the weekends.
-   Casual riders have a longer duration of rides than members do.
-   The average duration of rides is consistent throughout the week for members.
-   The average duration of rides is higher during the weekend for casual riders.

```{r export results, include=FALSE}
#Export summary file for further analysis or visualization
ridership_data <- all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n(),							#calculates the number of rides and average duration 
  average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)	
write.csv(ridership_data, file = 'avg_ride_length.csv')

ggsave("NumofRides.jpg",numrides)
ggsave("Duration.jpg",duration)
```

## Share

A presentation of this case study can be found [here](https://docs.google.com/presentation/d/1xu2mDvMNh751fhKryHo5MGF1VhV8aj-dEDGdl3CoWiM/edit?usp=sharing).